```{r ,echo=FALSE}
#want to ensure that we don't use scientific notation and display all the digits.
options(scipen=4, digits=8)
```



# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data
```{r}
data_all<-read.csv(unzip("activity.zip","activity.csv"))
data<-data_all[!is.na(data_all$steps),]
data$date <- as.Date(data$date)
```

## What is mean total number of steps taken per day?
```{r}
totalSteps <- tapply(data$steps,data$date,sum)
totalStepSummary <- summary(totalSteps)
hist(totalSteps,main="Histogram of Total Number of Steps per Day (ignoring NA)",xlab="Total number of Steps",breaks=20)
abline(v=totalStepSummary["Median"])
abline(v=totalStepSummary["Mean"])
totalStepSummary
```

The mean is `r totalStepSummary["Mean"]` and the median is `r totalStepSummary["Median"]`, they are displayed in the plot with vertical lines.

## What is the average daily activity pattern?
```{r}
averagePerInterval <- tapply(data$steps,data$interval,mean)
plot(names(averagePerInterval),averagePerInterval,type="l",)
max_ave <- names(averagePerInterval[averagePerInterval==max(averagePerInterval)])
abline(v=max_ave)
```

The index of the max ave is `r max_ave`.

## Imputing missing values
There are `r sum(is.na(data_all$steps))` missing values.
```{r}
data_all$averagePerInterval <- averagePerInterval[as.character(data_all$interval)]
data_all$steps <- ifelse(is.na(data_all$steps),data_all$averagePerInterval,data_all$steps)
totalStepsAll<-tapply(data_all$steps,data_all$date,sum)
hist(totalStepsAll,main="Histogram of Total Number of Steps per Day (Imputing NA)",xlab="Total number of Steps",breaks=20)
totalStepsAllSummary<-summary(totalStepsAll)
abline(v=totalStepsAllSummary["Median"])
abline(v=totalStepsAllSummary["Mean"])

totalStepsAllSummary
```

The mean is `r totalStepsAllSummary["Mean"]` and the median is `r totalStepsAllSummary["Median"]`, they are displayed in the plot with vertical lines. Since we are just adding more days with average results we make the histogram more peaked at the mean.


## Are there differences in activity patterns between weekdays and weekends?
```{r}
#create the weekend and weekday types
data_all$dayType <- ifelse(weekdays(as.Date(data_all$date)) %in% c("Saturday","Sunday"),"weekend","weekday")

#subset intervals by day type
weekendDays <- subset(data_all,subset=dayType=="weekend")
weekdayDays <- subset(data_all,subset=dayType=="weekday")

#take the mean of the steps over the intervals
aveWeekend <- with(weekendDays,tapply(steps,interval,mean))
aveWeekday <- with(weekdayDays,tapply(steps,interval,mean))

#build a data frame for each day type with the averages, want to set stringAsFactors false so we can turn the intervals into integers easily later  
weekdayDF<-data.frame(Interval=names(aveWeekday),Steps=aveWeekday,dayType="weekday",stringsAsFactors=F)
weekendDF<-data.frame(Interval=names(aveWeekend),Steps=aveWeekend,dayType="weekend",stringsAsFactors=F)

#concatenate the two data frames
allDF <-rbind(weekdayDF,weekendDF)

#do this so the x axis of the plot works right
allDF$Interval<-as.numeric(allDF$Interval)

#sort by the Interval
allDF<-allDF[order(allDF$Interval),]
library(lattice)
with(allDF,xyplot(Steps~Interval|dayType,type="l",layout=c(1,2),ylab="Number of Steps"))
weekendWeekdayPercentDiff<-(sum(aveWeekend)-sum(aveWeekday))/sum(aveWeekday)*100
```



Interesting differences between the work week activity and the weekend is that clearly thing get started later on the weekend. The steps increase sharply after 500 on the weekday where as on the weekend there's only a small peak at this point in the day. The smaller peak is very similiarly shaped, perhaps occational having to go into the office on the weekend? Having said that it does appear that once we get going there is a lot more walking done on the weekend than during the day. Summing up the average number of steps on the weekend vs weekday shows an increase of `r options(scipen=0, digits=2); weekendWeekdayPercentDiff`%. This would make sense as personal errands might involve more travel than a typical work day.